---
title: Alternative splicing analysis for single cell RNA sequencing
format:
  html:
    code-overflow: wrap
---
This vignette demonstrates alternative splicing analysis using gene, exon, and junction expression matrices generated by [PISA](anno.Rmd). Although most scRNA-seq protocols are 3' or 5' biased, partial exon coverage can still be captured. The spatial dissimilarity test in Yano exploits the **spatial distribution of features**, so **even lowly expressed features with strong spatial patterns** can be detected. By comparing exon/junction expression against their host gene's expression in spatial context, potential alternative splicing events are identified.

## 0. Prerequisite
- Make sure you have installed the R environment and the **latest version** of the `Yano` package (>= 1.2) before proceeding.
- The latest Yano uses **grid-based highlight rendering** for `TrackPlot`, enabling highlights that span continuously across multiple facet panels and tracks. If you have an older version installed, update with `devtools::install_github("shiquan/Yano")`.

## 1. Perform cell clustering with Seurat
```{r warning=FALSE}
require(Yano)
require(Seurat)
require(dplyr)
require(ggplot2)
require(Matrix)
```

```{r warning=FALSE}
# Read raw gene expression matrix
exp <- ReadPISA("./exp/")
```
In this section, we will perform the standard Seurat analysis pipeline. 


*Note:* The spatial dissimilarity test does not depend on clustering. Changing resolution or other parameters for `FindClusters` and `RunUMAP` has no impact on the results.
```{r warning=FALSE}
# Create Seurat object and filter droplets with fewer than 1000 genes
obj <- CreateSeuratObject(exp, min.features = 1000, min.cells = 10)

# Filter low quality droplets
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
obj <- subset(obj, nFeature_RNA < 9000 & percent.mt < 20)

# Downsampling to 2000 cells for fast testing
obj <- obj[, sample(colnames(obj),2000)]

# We run the cell clustering analysis with Seurat pipeline
obj <- NormalizeData(obj) %>% FindVariableFeatures() %>% ScaleData() %>%  RunPCA(verbose=FALSE) %>% FindNeighbors(dims = 1:10, verbose=FALSE) %>% FindClusters(resolution = 0.5, verbose=FALSE) %>% RunUMAP(dims=1:10, verbose=FALSE)

DimPlot(obj, label=TRUE, label.size = 5, label.box = TRUE)
```

## 2. Perform alternative splicing with exon assay
Here we compare exon expression with their host gene's expression in a spatial context. "Spatial" refers to cell organization: in this vignette we use PCA space, but the method also supports spatial coordinates, lineage trajectories, or integration spaces such as harmony. The workflow has three steps. First, load exon data as a new assay. Second, run spatial autocorrelation to select informative exons. Third, define exon–gene binding relationships and run the spatial dissimilarity test to obtain p-values and adjusted p-values.

```{r warning=FALSE, fig.width=14, fig.height=6}
# Read exon count matrix file
exon <- ReadPISA("./exon/")

# Load the exon expression to Seurat object as a new assay, make sure the exon matrix has the same cells.
obj[['exon']] <- CreateAssayObject(exon[, colnames(obj)], min.cells=20)

# Switch work assay to exon
DefaultAssay(obj) <- "exon"
# Empty info for exon features
head(obj[['exon']][[]])

obj <- ParseExonName(obj)
# Gene name and location are parsed from exon name
head(obj[['exon']][[]]) %>% DT::datatable()

# Normalize the data for spatial autocorrelation test
obj <- NormalizeData(obj)
obj <- RunAutoCorr(obj)
```

The permutation test is computationally expensive. Below we use `perm=20` for quick results; the default is 100 permutations for better accuracy. If you are running Yano for the first time on a new dataset, `perm=20` saves time and provides a useful overview.
```{r warning=FALSE, fig.width=14, fig.height=6}
obj <- RunSDT(obj, bind.name = "gene_name", bind.assay = "RNA", perm=20)

# Now p values and adjusted p values have been generated
head(obj[['exon']][[]]) %>% DT::datatable()

# Plot feature binding test plot
FbtPlot(obj, val = "gene_name.padj")
```

The chromosome names are long and often overlap in the plot. You can resize labels or remove the "chr" prefix. The Y chromosome and mitochondrial genes can also be excluded.
```{r warning=FALSE, fig.width=14, fig.height=6}
sel.chrs <- c(1:21, "X")
FbtPlot(obj, val = "gene_name.padj", remove.chr = TRUE, sel.chrs = sel.chrs)

# Let's see how many exons are expressed in different spatial pattern with their genes
obj[['exon']][[]] %>% filter(gene_name.padj < 1e-5) %>% DT::datatable() 

# Random select a gene and its exons and visulize with FeaturePlot.
FeaturePlot(obj, features = c("chr19:16095264-16095357/+/TPM4", "TPM4"), ncol=2)

# The default color and parameters perhaps not easily to tell the difference between exon and its binding gene expression. Let's change the scaled colors and enlarge point size and order by expression.
require(RColorBrewer)
FeaturePlot(obj, features = c("chr19:16095264-16095357/+/TPM4", "TPM4"), ncol=2, order = TRUE, pt.size=1) & scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu")))
```
The `RatioPlot` function maps the ratio of exon expression to gene expression on the UMAP. TPM4 is lowly expressed in some cell groups, while the exon chr19:16095264-16095357/+/TPM4 shows a higher ratio in those same groups.
```{r}
RatioPlot(obj, features = c("chr19:16095264-16095357/+/TPM4"), assay = 'exon', bind.assay = 'RNA', bind.name = "gene_name", order = TRUE, pt.size=1)
```

In the plots above, the inconsistency between exon and gene expression may indicate differential exon usage. Next we generate a track plot to inspect read coverage for the exon and gene body.

Yano loads gene locations from a GTF file rather than Bioconductor databases (e.g., `org.Hs.eg.db`). This avoids inconsistencies from differing annotation versions across sources. We recommend using the same GTF file throughout your project. The `gtf2db` function loads a GTF file into memory.
```{r}
gtf <- gtf2db("./gencode.v44.annotation.gtf.gz")
```

The track plot shows read coverage per cell group, specified by `cell.group`. Unlike IGV which uses read depth, this plot shows UMI depth. The cell barcode and UMI tags default to "CB" and "UB" (`cell.tag` and `umi.tag`). For each cell group, UMI depth is normalized by the number of cells, so the tracks are directly comparable across groups. If `cell.group` is not set, raw UMI depth per location is shown.

Highlighted regions are rendered with **dark orange boundary lines** and a semi-transparent grey fill. In the latest Yano, highlights use grid-based rendering that **spans continuously across all facet panels and tracks** (coverage, BED features, and gene annotations) with a small visual protrusion, making them clearly distinguishable from the panel data. Overlapping highlights produce additive darkening.

```{r message=FALSE, fig.width=12, fig.height=8}
TrackPlot(bamfile = "Parent_SC3v3_Human_Glioblastoma_possorted_genome_bam.bam", gtf = gtf, gene = "TPM4", cell.group = Idents(obj), highlights = c(16095264,16095357) )
```

## 3. Load junction assay
Junction expression provides a complementary view of alternative splicing. Junctions capture reads spanning adjacent exons. They are named similarly to exons but with different coordinates: the junction start is the end of the upstream exon, and the junction end is the start of the downstream exon.
```{r warning=FALSE, fig.width=14, fig.height=6}
junction <- ReadPISA("./junction/")
obj[['junction']] <- CreateAssayObject(junction[, colnames(obj)], min.cells=20)

DefaultAssay(obj) <- "junction"
obj <- NormalizeData(obj)

# select spatial autocorrelated junctions
obj <- RunAutoCorr(obj)

# Parse the gene name and coordinates from junction names
obj <- ParseExonName(obj)

# perform dissimilarity test between junctions and their binding genes
obj <- RunSDT(obj, bind.name = "gene_name", bind.assay = "RNA", perm=20)

FbtPlot(obj, val="gene_name.padj", remove.chr=TRUE, sel.chrs = sel.chrs)

obj[['junction']][[]] %>% filter(gene_name.padj<1e-5) %>% DT::datatable()

# Because both exon and junction are compared with gene, so it's reasonable to combine these two assays in one plot
FbtPlot(obj, val="gene_name.padj", assay = c("exon", "junction"), col.by = "assay", shape.by = "assay", pt.size = 2, remove.chr = TRUE, sel.chrs = sel.chrs, cols = c("red", "blue"))

# We can find there is an exon and a junction at chromosome 12 with very low p value (<1e-5), let's see which gene they are located
obj[['exon']][[]] %>% filter(chr == "chr12" & gene_name.padj < 1e-5) %>% DT::datatable()
obj[['junction']][[]] %>% filter(chr == "chr12" & gene_name.padj < 1e-5) %>% DT::datatable()

FeaturePlot(obj, features = c("chr12:56161387-56161465/+/MYL6","chr12:56160320-56161387/+/MYL6", "MYL6"), order = TRUE, pt.size = 2, ncol=3) & scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu")))

# We could also plot the expression ratio of these exons or junctions on umap
p1 <- RatioPlot(obj, assay = "exon", bind.assay = "RNA", bind.name = "gene_name", features = "chr12:56161387-56161465/+/MYL6")
p2 <- RatioPlot(obj, assay = "exon", bind.assay = "RNA", bind.name = "gene_name", features = "chr12:56160626-56160670/+/MYL6")
p3 <- RatioPlot(obj, assay = "junction", bind.assay = "RNA", bind.name = "gene_name", features = "chr12:56160320-56161387/+/MYL6")
cowplot::plot_grid(p1,p2,p3, ncol=3)
```

We then visualize the track plot for this gene, including junction reads by setting `junc=TRUE`. Two adjacent highlight regions are specified as a list, producing **overlapping highlights** — the overlapping portion appears darker due to additive alpha blending, and each highlight is marked with a dark orange boundary line at its start position.

```{r message=FALSE, fig.width=12, fig.height=8}
TrackPlot(bamfile = "Parent_SC3v3_Human_Glioblastoma_possorted_genome_bam.bam", gtf = gtf, gene = "MYL6", cell.group = Idents(obj), junc = TRUE, highlights = list(c(56160320,56161387),c(56161387,56161465)))
```
The exon chr12:56161387-56161465/+/MYL6 appears highly expressed in group 3 in the track plot, yet its expression in the feature plot is lower than expected. This is because the exon overlaps with other exons from different transcripts. Only reads fully contained within the exon are counted toward its expression; partially overlapping reads are excluded.

The overlapping exon chr12:56161387-56161575/+/MYL6, by contrast, shows strong expression in group 3. Note that when a read is fully contained within multiple overlapping exons, PISA counts it for all of them. See PISA’s manual for details.

```{r fig.width=10, fig.height=5}
p1 <- DimPlot(obj, label=TRUE, label.size = 5, label.box = TRUE)
p2 <- FeaturePlot(obj, features = c("chr12:56161387-56161575/+/MYL6"), order = TRUE, pt.size = 1) & scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu")))
p1 + p2
```

## 4. Heatmap analysis for highlight group specific alternative splicing
The spatial dissimilarity test identifies alternatively spliced exons and junctions across all cells, but does not pinpoint which cell groups drive the signal. To address this, we extract scaled expression for the selected exons and their host genes, then perform co-clustering. The heatmap below uses [ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html) to visualize the distribution across cell groups.

```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
obj[['exon']][[]] %>% filter(gene_name.padj<1e-5) %>% rownames -> exons
obj[['exon']][[]] %>% filter(gene_name.padj<1e-5) %>% pull(gene_name) -> bind.genes
DefaultAssay(obj) <- "RNA"
obj <- ScaleData(obj, features = unique(bind.genes))
DefaultAssay(obj) <- "exon"
obj <- ScaleData(obj, features = exons)

dat1 <- GetAssayData(obj, assay = 'exon', layer = 'scale.data')
dat2 <- GetAssayData(obj, assay = 'RNA', layer = 'scale.data')
idents <- sort(Idents(obj))
order.cells <- names(idents)

dat2 <- dat2[bind.genes,]
rownames(dat2) <- exons

dat <- cbind(dat1, dat2)

require(ComplexHeatmap)
d <- dist(dat)
hc <- hclust(d)
idx <- hc$labels[hc$order]

ha <- HeatmapAnnotation(group=idents, border = TRUE)
ht1 <- Heatmap(dat1[idx, order.cells], cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = FALSE, border = TRUE,  top_annotation = ha, name = "exon", column_title = "exon")
ht2 <- Heatmap(dat2[idx, order.cells], cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = FALSE, border = TRUE,  top_annotation = ha, name = "gene", column_title = "gene", row_names_max_width = max_text_width(rownames(dat2), gp = gpar(fontsize = 12)))

ht <- ht1 + ht2
draw(ht, heatmap_legend_side = "left",  annotation_legend_side = "left")
```


## 5. Test between exon and exon skipped reads

The spatial dissimilarity test is not limited to gene-level binding features. Here we test exon expression against reads that skip the exon (exclude assay). This is analogous to the Percent Spliced In (PSI) method widely used in bulk and single-cell RNA-seq: `PSI = exon reads / (exon reads + reads skipping this exon)`.

```{r  message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
# The reads that skip exons are annotated using the `-psi` option in PISA anno, and these counts are stored in the `exclude` directory. We then load these excluded counts into a new assay.
exclude <- ReadPISA("exclude/")
obj[['exclude']] <- CreateAssayObject(exclude[,colnames(obj)], min.cells = 10)
DefaultAssay(obj) <- "exclude"
# Normalize counts for exon-excluded reads
obj <- NormalizeData(obj)
# Then we switch to exon assay
DefaultAssay(obj) <- "exon"

# Because the feature names in the exclude assay are exactly the same as those in the exon assay, they represent the reads that skip each corresponding exon. Therefore, we set up the binding feature using the exon name itself.
obj[['exon']][['exon_name']] <- rownames(obj)
obj[['exon']][['exon_name']] %>% head

# Then we perform spatial dissimilarity test between exon and exclude, mode 1
obj <- RunSDT(obj, bind.name = "exon_name", bind.assay = "exclude")

# Swith to exon exluded assay
DefaultAssay(obj) <- "exclude"
obj <- RunAutoCorr(obj)
obj <- ParseExonName(obj)
obj[['exclude']][['exon_name']] <- rownames(obj)

obj <- RunSDT(obj, bind.name = "exon_name", bind.assay = "exon")

FbtPlot(obj, val = "exon_name.padj", remove.chr = TRUE, assay = c("exclude", "exon"), shape.by = "assay", col.by = "assay", cols = c("yellow", "green"), pt.size = 2)
```
```{r  message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
# Let's how many exons can be prioritized by both exon assay and exclude assay
obj[['exclude']][[]] %>% filter(exon_name.padj<1e-5) %>% rownames -> sel1
obj[['exon']][[]] %>% filter(exon_name.padj<1e-5) %>% rownames -> sel2
intersect(sel1,sel2)

DefaultAssay(obj) <- "exclude"
p1 <- FeaturePlot(obj, features = c("chr19:16095264-16095357/+/TPM4"),order = TRUE)
DefaultAssay(obj) <- "exon"
p2 <- FeaturePlot(obj, features = c("chr19:16095264-16095357/+/TPM4"),order = TRUE)
p3 <- PSIPlot(obj, exon.assay = "exon", exclude.assay = "exclude", features = c("chr19:16095264-16095357/+/TPM4"),order = TRUE)
cowplot::plot_grid(p1,p2,p3, ncol=3)
```

Here, we set `max.depth = 1` caps the depth and setting `display.genes = "TPM4"` filters to the gene of interest, making the highlighted exon more visible.
```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
TrackPlot(bamfile = "Parent_SC3v3_Human_Glioblastoma_possorted_genome_bam.bam", gtf = gtf, gene = "TPM4", cell.group = Idents(obj), highlights = c(16095264,16095357), junc = TRUE, max.depth = 1, display.genes = "TPM4")
```


Exon or junction expression is part of gene expression, so inverse expression patterns can indicate alternative splicing. By contrast, exon-skipped reads are largely independent of exon expression, offering higher sensitivity for detecting splicing events.

The spatial dissimilarity test does not model the spatial dependency of the binding feature, so many events may be prioritized when the binding feature is sparse. Intersecting results from both directions (exon vs. exclude and exclude vs. exon) helps reduce false positives from low coverage.

Mode 3 specifically detects strong inverse expression patterns by summing exon reads and exon-excluded reads as the binding assay. Events found by mode 3 are a subset of those found by mode 1, making it a stricter filter.

```{r  message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
DefaultAssay(obj) <- "exclude"
obj <- RunSDT(obj, bind.name = "exon_name", bind.assay = "exon", mode = 3, prefix = "mode3")
DefaultAssay(obj) <- "exon"
obj <- RunSDT(obj, bind.name = "exon_name", bind.assay = "exclude", mode = 3, prefix = "mode3")
FbtPlot(obj, val = "mode3.padj", remove.chr = TRUE, assay = c("exclude", "exon"), shape.by = "assay", col.by = "assay", pt.size = 2, cols=c("yellow", "green"))
```


In this case study, we applied the spatial dissimilarity test to multiple feature pairs. The method works on the entire cell population without requiring prior clustering or annotation, making it suitable for exploratory analysis. We recommend testing both exons and junctions against their host genes, particularly in 3’ or 5’ biased scRNA-seq data. The exon vs. exon-skipped comparison provides additional power to detect exon exclusion events. For cell-cluster-specific patterns, follow up with heatmaps and co-clustering.

## Questions?
If you have any questions regarding this vignette and the usage of Yano, please feel free to report them through the [discussion forum](https://github.com/shiquan/Yano/discussions). When submitting your query, please ensure you attach the commands you used for better clarity and support.

```{r}
Command(obj)

sessionInfo()
```

    

<div id="page-nav-container"></div>
<script src="navigation.js"></script>
