---
title: Alelle specific gene expression analysis for single cell RNA sequencing
format:
  html:
    code-overflow: wrap
---
This vignette demonstrates allele-specific gene expression analysis using gene expression and single nucleotide variant (SNV) data generated by [PISA](anno.Rmd).


## 0. Perform cell clustering with Seurat

We begin with cell clustering based on gene expression, identical to the first step in the [Alternative splicing analysis vignette](Yano_AS.Rmd). Skip this section if you are continuing from there.

```{r warning=FALSE, fig.align='center'}
require(Seurat)
require(ggplot2)
require(dplyr)
require(Yano)

exp <- ReadPISA("./exp/")

obj <- CreateSeuratObject(exp, min.features = 1000, min.cells = 10)
obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
obj <- subset(obj, nFeature_RNA < 9000 & percent.mt < 20)

# Downsampling to 2000 cells for fast testing
obj <- obj[, sample(colnames(obj),2000)]

# To improve visualization, we reduced the resolution when identifying cell clusters.
obj <- NormalizeData(obj) %>% FindVariableFeatures() %>% ScaleData() %>%  RunPCA(verbose=FALSE) %>% FindNeighbors(dims = 1:10, verbose=FALSE) %>% FindClusters(resolution = 0.1, verbose=FALSE) %>% RunUMAP(dims=1:10, verbose=FALSE)
DimPlot(obj, label=TRUE, label.size = 5, label.box = TRUE)
```
## 1. Perform allele-specific gene expression analysis by using SNV anchors

In this vignette, "genetic variant" refers to all alleles at a given locus, including both the reference allele and any alternative alleles. All are annotated and counted by PISA.

```{r message=FALSE}
var <- ReadPISA("./var")
obj[['var']] <- CreateAssayObject(var[,colnames(obj)], min.cells = 10)
# Switch working assay to 'var'
DefaultAssay(obj) <- "var"

obj <- NormalizeData(obj)

# Skip loading the GTF if you already have it in memory
gtf <- gtf2db("./gencode.v44.annotation.gtf.gz")

# Currently the meta information for the variant assay is empty
obj[['var']][[]] %>% head

obj <- annoVAR(obj, gtf = gtf)

# After annotation, locations are parsed from feature names and overlapping genes are annotated.
obj[['var']][[]] %>% head %>% DT::datatable()

# Select spatially autocorrelated genetic variants
obj <- RunAutoCorr(obj)
```

For allele-specific expression, we use mode 2 instead of the default mode 1. Mode 2 compares a variant's expression against other variants at the same locus. Since no separate binding assay is specified, counts at each locus are aggregated, then the test feature is subtracted from the locus total.

```{r fig.width=12, fig.height=5, fig.align='center'}
obj <- RunSDT(object = obj, bind.name = "locus", mode = 2)

sel.chrs <- c(1:21, "X")
FbtPlot(obj, val = "locus.padj", remove.chr = TRUE, sel.chrs = sel.chrs)

# Randomly pick a pair of features at the same locus
obj[['var']][[]] %>% filter(locus.padj < 1e-10) %>% DT::datatable()
FeaturePlot(obj, features = c("chr10:17237326C>T/+", "chr10:17237326C=/+", "VIM" ), order=TRUE, pt.size = 1, ncol=3)

# Zoom into a specific region
FbtPlot(obj, val = "locus.padj", chr="chr10", start = 17225000, end = 17240000, gtf = gtf)
```
The T allele of VIM is expressed across most cell groups, while the C allele is primarily restricted to groups 1 and 4.

The spatial dissimilarity test identifies many cases of allele-specific expression. This phenomenon can be driven by gene imprinting, genetic recombination, somatic variants, and other mechanisms. For detailed case studies, see our manuscript.



```{r}
# Save the Seurat object for further use
saveRDS(obj, file = "glioblastoma5k.rds")
```

```{r}
Command(obj)

sessionInfo()
```

<div id="page-nav-container"></div>
<script src="navigation.js"></script>
